Symbol Recognition using Particle Swarm Algorithms

### 2008-11-23 trunk/writing/outline.tex
the paper before i edit after dina requested changes 2

\documentclass[a4paper,10pt]{IEEEconf}
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
%\usepackage[T1]{fontenc}
\usepackage[dvips,pdftex]{graphicx}
%\usepackage{multicol}
\usepackage{subfigure}
\usepackage{named}
%\usepackage{fancybox}
%\usepackage{natbib}
\bibliographystyle{named}
\usepackage[dvips,pdftex]{hyperref}
%opening
\title{ Symbol Recognition using Particle Swarm Algorithms}
%\title{A Study on Using Particle Swarm Algorithms in Symbol Recognition }
\author{Maha El Meseery,  Mahmoud Fakhr El Din,  Samyah Mashaly,
\\  Magdah Fayek,  Nevin Darwish}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying the symbols the user draws using a single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes into objects that can be easily manipulated. This paper uses Particle Swarm Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which will be further identified. The results showed that using PSO improves segmentation results which guide the symbol recognition phase. Final recognition has improved accuracy after the PSO algorithm using a SVM classifier.  
 %Hand drawn sketches are the natural method of exchanging thoughts between engineers and humans in general. Nevertheless, researchers barely give attention to improve current gesture and sketch recognition systems. In a sketch understanding system, users need to feel the freedom of paper and pencil along with the flexibility of computer manipulation. 
\end{abstract}
\section{Introduction}
%A picture speak a thousand words is also true for sketches. 
Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Currently, engineers use paper and pencil design in early stages of design. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketchs more effectively than with only papers.  Increasing the interaction between the computer and the users in sketch and CAD systems has been recently the reason for the emerging of few advanced sketch recognition systems. Sketch recognition system converts the drawings and informal graphs the user draws into objects and symbols can be used in further interaction with the system.

Sketch recognition researchers divide the system into four main steps preprocessing, segmentation, symbol recognition and finally sketch understanding. The preprocessing process saves the location of points from the hardware and collects basic information about the stroke data then proceed to remove noise and primary data calculation.  In segmentation step, the system  divide the stroke into a set of simple geometrical primitives or segments.  The third step of the system cluster strokes and segments to formulate symbols that can be recognized by a classifier system. Finally, the sketch understanding step uses priori information about the sketch to support identification of symbols by excluding any possible invalid symbols. 
%the basic symbols are identified in consistent with the whole sketch. \\
%from the user information from the users converts the strokes the user draw into segments which is recognized into basic symbols that can be easily manipulated. Finally, the basic symbols are interpreted in consistent with the whole sketch. \\

Section \ref{Sec:RelatedWork} provides brief description to related work. The remaining of this papers will be as following: Section \ref{PSO} will explain the general particle swarm algorithm. Section \ref{Sysdisc} will describe the system in details.  Section \ref{sec:Experiments} will describe the experiments that was preformed. Finally, section \ref{ConclusionandFutureWork} will state the conculusion and future work. 
%The next section will relate previous work in the field. The remaining of the paper will be as following; Section \ref{PSO} will explain the general particle swarm algorithm. Section \ref{Sysdisc} will describe the system block diagram and then section \ref{Prepross} will explain the preprocessing step, section \ref{seg} will describe the segmentation algorithm in details. Finally, section \ref{sec:Experiments} will describe the experiments that was preformed.\\
\section{Literature Review}
\label{Sec:RelatedWork}
%%%What is sketch recognitions. Why we need sketch recognition systems. 
The field of symbol recognition has gained interest in the last few years. A wide variety of techniques were used either on segmentation process or symbol recognition.

%To segment strokes
In order to segment strokes, \cite{meanshift10} introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on this area. %\citeauthor{HeloiseBeautification} used a dynamic programming algorithm to solve the segmentation problem \cite{HeloiseBeautification}.

Complex and hybrid algorithm were also introduced as in  \cite{earlyprocess} where they generated different sets of segmentation based on both curvature and speed dominant points then followed by a choice a segmentation with the least error from the hybrid set. Additionally, a dynamic programming algorithm was used to obtain optimum segmentation of the input strokes in \cite{GeometryAndDomain102}. Genetic algorithm was also used by \cite{CruveDivisionSwarm} to optimally divide digital curves into lines and curves. \citeauthor{PolygonApproximationPSO} used PSO to convert digital curves into polygons\cite{PolygonApproximationPSO}.

Many different approaches have been investigated to achieve final recognition of symbols. Graph searching and template matching where used in \cite{mulitstroke5,GraphBased07}, where a graph of the drawn symbol is generated and matched with a stored set of graphs templates to reach a classification decission. Geometrical and spatial descriptive language was used in \cite{SketchRead2007,Ladder30} where an architecture simillar to compiler was used to recognize symbols. \citeauthor{imagetrainable48} used trainable image template to classify different set of symbols \cite{imagetrainable48}. The use of SVM and HMM classification was implemented in\cite{SVM300,SmartSketch56,HMM53}.
% Other techniques as  [          ] was used  in []. \\  
 %In the next section the particle swarm algorithm will be explained. After that the system block description and detailed algorithms will be described. We will then proceed to describe the experiments preformed and the achieved results.  
%\subsection{Sketch Recognitions systems}
%Details of Sketch systems. 
%Related work and research in this area. 
%Writing on the pros and cons of current systems.  
%Write on the difficulties of the sketch recognition systems. 
  %Some tried to detect dominant or corner points based on data generated from speed   \cite{earlyprocess} and mostly used curvature of the stroke drawn \cite{earlyprocess,meanshift10}. 
%Particle swarm algorithms was used in \cite{} to fit a set of points into ellipse or higher order curves.
%The first set of features introduced to be used on gesture recognition was introduced by Rubine in \cite{gestureexample12}. later,  Spatial relations and features where used in  \cite{cognitivesketch18,SRGraph57}  . \citeauthor{GeometryAndDomain102} introduced ink density feature which they used.  The use of Zernike moments  was used  as the only features to the SVM classifiers in \cite{HeloiseBeautification}.  \cite{SketchRead2007} used a set of features varied from relative spatial features and some compositional features.  \\
%After segmentation researchers varies on the methods in which to continue.
\section{Particle Swarm Algorithm}
\label{PSO}
%What is particle swarm algorithm and how it was used in related researches. 
The main idea of  Particle Swarm Algorithm (PSO) is to represent each agent with a particle from the solution space. Each agent moves the particle with a direction and velocity based on the following equation [\ref{eq:Swarm}].
\begin{equation}
%\[
v_{ij}  = v_{ij}  + c_1 r_1 (lbest_{ij}  - p_{ij} ) + c_2 r_2 (gbest_{ij}  - p_{ij} )
%\]
\label{eq:Swarm}
\end{equation}
\begin{equation}
%\[
p_{ij}=p_{ij}+v_{ij},
%\]
\end{equation}
 Equation [\ref{eq:Swarm}] shows how velocity and direction of each particle is computed where $r_1$ \& $r_2$ are random variable and $c_1$ \& $c_2$ are the swarm system variables. Each iteration the global best $g_{best}$ particle and  the agent local best $l_{best}$ particle are evaluated based on the maximum fitness functions of  all particles. The solution is found after a maximum number of iteration or after an error threshold is achieved.
 \begin{equation}
   P(i)\Leftarrow 
\{
\begin{array}{c} 
1 \quad \quad if\quad r_{3}>t  \\

0 \quad \quad if\quad r_{3}<t 
\label{eq:descrite}
\end{array}\}
\end{equation}
  Equation \ref{eq:descrite} where t is a threshold and $r_{3}$  is a random variable, is used to change general swarm algorithm into binary particle which will handle particle values of either 0 or 1. .  
\section{System Description}
\label{Sysdisc}
%The stroke is the path of points from the instant the pen is down till it is up.
Figure. \ref{fig:Blockdiagram} shows that the main system blocks are 1) preprocessing, 2) segmentation, 3) clustering and feature extraction, and finally 4) classification.

 %Due to sloppiness of users and hardware glitches the system preprocesses the data before further processing. 
The preprocessing is responsible of capturing the input data and removing the noise from it. The system then proceeds to estimate a set of possible dominant points to help in the segmentation process. % The time differences, direction, speed and curvature of all points of the stroke is computed along with some other geometrical data. The system then proceed to estimate a set of possible dominant points to help in segmentation process.  \\
  %then passes through the segmentation process which will try to

@@

 %the stroke first If the ellipse detection fails the stroke is passed to the segmentation algorithms which will pass it to the two PSO algorithms described below the segmentation with the minimum error will be the chosen segmentation.  The segmentation is then added to the set of un-recognized segments in the system. \\% This part is repeated for each stroke. 
 The clustering alogrithms starts to group segments togther after the segmentation step. The system let the user draws the symbol by using any number of strokes, a set of unrecognized segments is passed to the clustering algoirthm to generate a symbol and compute a feature vector for it.  The final step is a SVM classifier that will use the features computed to classify the segments into one of the previously trained classes. %  Or to determine the symbol of the given segments from the set of preciously trained symbols.

 %with a symbol from the training set.% The system compute composite set of features some are statistical other are spatial features based on the type primitives. Finally, the strokes are classified into the corresponding classes.% the classifier identifies the strokes into a symbol from the set of known symbols.\\
\begin{figure}[]
	\centering
		\includegraphics[scale=0.35]{images/Blockdiagram.pdf}
	\caption[  ]{The system Block Diagram}
	\label{fig:Blockdiagram}
\end{figure}
\subsection{Preprocessing}
\label{Prepross}
%Some data is extracted from the points in the stroke.
 It is noted that as the user draws a shape the pen will slow down near corners and picks up speed when drawing a straight lines. Therefore, the speed and time difference data were widely used in sketch understanding systems \cite{earlyprocess}.  The direction and curvature data are used to determine the points with high angular changes along the path of points \cite{meanshift10}.

@@

 %system compute speed, distance data for all points in the stroke. Curvature is computed using estimation used in [] where direction computed is angle between two lines. Curvature is 
 %$\Delta d/  \Delta S $ where d is the difference direction of point and s is difference in distance between points. 
 All these calculations are performed in real time while the user draws the strokes.
 The complexity of compution is $O(n)$ where $n$ is number of points. \\
 Figure \ref{fig:speed2} shows the computed data for the stroke drawn in fig. \ref{fig:orignalStroke}.  If you look at the data curves you will see that dominant points are characterized by lower speed values and higher curvature and direction values. \\% the lower speed points correspond to vertex and dominant point's location. The higher the direction and curvature data correspond to location with higher change in curvature witch promote the location for vertices. \\
After the system computes all the speed, time difference and curvature data it proceeds to detect the points with low velocity and high curvatures. Using simple differentiations to detect local extreme points resulted in false points due to the non smooth curves. So instead the following procedure was used, firstly the mean of the curve is used as threshold for lower values then the high values are further investigated to find the extreme points. Hence, the mean of the direction curve is calculated and used as threshold. This threshold is used to separate the curve into regions; each region is defined as the part between two intersection points of the threshold line and the curve. Only regions higher than the threshold value are processed when computing the maximum of each region. The point that corresponds to those maximum values are labeled possible dominant points.   \\
   %If we tried to differentiate the curve the result will be false threshold values we divide the data into regions of data higher or lower that the threshold. This will let us only look for high data only. The maximum of each region is then computed and reported as a possible vertex point. 
\begin{figure}[]
	\centering
		\includegraphics[scale=0.3]{images/orignalStroke.pdf}
	\caption{Stroke Sample}
	\label{fig:orignalStroke}
\end{figure}
The system repeat this process to curvature, time difference and speed curves. All the points labeled as possible dominant points are saved into a single array. They are used as input and guiding data for the next stage of segmentation. 
\begin{figure*}[]
\begin{minipage}[b]{0.5\linewidth}
	\centering

		\subfigure[ The Speed of data] {\includegraphics[scale=0.15]{images/speed2.pdf}}

		\subfigure[ Direction ] {\includegraphics[scale=0.15]{images/direction2.pdf}}
					\end{minipage}
					\begin{minipage}[b]{0.5\linewidth}
			\subfigure[Curvature] {\includegraphics[scale=0.15]{images/curvature2.pdf}}

			\subfigure[ Time Difference] {\includegraphics[scale=0.15]{images/time2.pdf}}
			\end{minipage}

	\caption[Data Curves]%{The data of the stroke}

@@

After computing the primary data the system now tries to segment the stroke into a set of primitives. The segmentation algorithm first tries to detect if the stroke is an ellipse. If the stroke proved to be an ellipse then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed into two particle swarm algorithms that will divide the stroke to either lines or lines and curves. The algorithms takes the stroke points along with the possible dominant points computed then produce a set of dominant points which are connected with either lines or curves (see fig. \ref{fig:Blockdiagram}).  The next section will describe the ellipse detection algorithm and both the two particle swarm algorithms used to divide the stroke. \\

@@

\subsubsection{ Ellipse Detection } 

@@

The process starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the center of the ellipse. The axes of the ellipse are estimated as the w/2 and h/2 of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation (see \ref{eq:circleFit} where N is number of points in the stroke, $a,b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point, $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke). After only few loops the error and a confidence values  are computed to check if the storke can be labled as ellipse or not.\\ %is computed to check if stroke is an ellipse. \\

@@

\end{equation}
\subsubsection{ Discrete particle swarm algorithm}

@@

\subsubsection{ problem definition}

@@

let's define the arc $\widehat{x_ix_j}$ as the consecutive set of points from point $x_i,x_{i+1} \cdots,x_j$. The line
$\overline{x_i x_j} $ as the straight line connecting point $x_i$ to point $x_j$. The approximation error is computed by the equation \ref{eq:ErrorSwarm1}  where M is the number of dominant points in this solution.  The error $ e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$.  \\

@@

\end{equation}

@@

\end{equation}%\]
The fitness is computed using equation \ref{eq:fitnessSwarm1} where N is the number of points in the stroke, D is the number of point in the solution that was previously labeled as a possible dominant point and E is the computed error and $\varepsilon$ is the error threshold.   As you can see when the error is larger than the threshold is given a -ve value to lower the value of solution otherwise the system will favor the lower number of vertices.\\ %we will want to lower the number of vertices. \\

@@

%Alg2: 
The second algorithm has the same problem formulation but different fitness and error functions. It was previously introduced in  \cite{CruveDivisionSwarm} where but genetic programming was used as the optimizing algorithm.  The error of both circle and line estimation are computed for each segment, the approximation with the lower error value will be the chosen approximation of this segment. The sum of the approximation error of all segments is the total error of the particle.  The error is computed by equation \ref{eq:errorSwarm2} where M is the number of segments in the solution, $D_i$ is the minimum approximation error of curve and line approximations $min(d_c,d_l)$ as computed by \cite{CruveDivisionSwarm}.  The fitness is computed by the equation \ref{eq:fitnessSwarm2} where E is the error and M is number of segments and k is a parameter tweaked to get minimum number of segments. \\

@@

\end{equation}

@@

\end{equation}

@@

\cite{PolygonApproximationPSO} used a merge and divide algorithm after each loop of the swarm system to refine the solution but we used another enhancement method. After each loop in the swarm algorithm, each particle loops on the set of selected dominant points to enhance the solution. Each dominant point is checked to find if it was labeled before as a possible dominant (computed as in section \ref{Prepross}), if not the point is moved to the nearest labeled point.    

@@

After the user draw all strokes of the symbol he has to wait 10 seconds or press finish button beside the drawing area. The set of unrecognized strokes is grouped together along with their segmentation as input to the feature extraction process. The system uses a composite set of features include Rubine feature set \cite{gestureexample12}, Zernike moments \cite{HeloiseBeautification}, ink density \cite{GeometryAndDomain102} and some structural information like number of perpendiculars lines , number of parallel lines and types of primitives in each symbol. After computing the features the symbol is introduced to the classifier. The system used a support vector machine (SVM) classifier with Gaussian kernel (RBF kernel) \cite{libsvm}.\\%Some features like number of parallel, perpendicular and intersecting lines are computed from the segmentation other like density, centroid and Zernike moments are computed from the raw stroke points. 

>>>>

\documentclass[a4paper,10pt]{IEEEconf}
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
%\usepackage[T1]{fontenc}
\usepackage[dvips,pdftex]{graphicx}
%\usepackage{multicol}
\usepackage{subfigure}
\usepackage{named}
%\usepackage{fancybox}
%\usepackage{natbib}
\bibliographystyle{named}
\usepackage[dvips,pdftex]{hyperref}
%opening
\title{Symbol Recognition using Particle Swarm Algorithms}
%\title{A Study on Using Particle Swarm Algorithms in Symbol Recognition }
\author{Maha El Meseery,  Mahmoud Fakhr El Din,  Samyah Mashaly,
\\  Magdah Fayek,  Nevin Darwish}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system should immediately interprets their strokes into objects that can be easily manipulated. This paper uses Particle Swarm Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results shows that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses SVM classifier which  further improves the final recognition accuracy.  %u Final recognition has improved accuracy after the PSO algorithm using a SVM classifier.  
 %Hand drawn sketches are the natural method of exchanging thoughts between engineers and humans in general. Nevertheless, researchers barely give attention to improve current gesture and sketch recognition systems. In a sketch understanding system, users need to feel the freedom of paper and pencil along with the flexibility of computer manipulation. 
\end{abstract}
\section{Introduction}
%A picture speak a thousand words is also true for sketches. 
Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Engineers use paper and pencil design in early stages of design. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers.  Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of few advanced sketch recognition systems. Sketch recognition system converts the drawings and informal graphs user draws into objects and symbols that can be used in further interaction with the system.

Sketch recognition is divided into four main steps preprocessing, segmentation, symbol recognition and finally sketch understanding. The preprocessing process saves user input stroke points from the hardware and collects basic information about the stroke then proceeds to remove noise and primary data calculation.  In segmentation phase, the system  divides the stroke into a set of simple geometrical primitives or segments.  The  third phase, sketch recognition, is clustering strokes and segments to formulate symbols that can be recognized by a classifier system. Finally, the sketch understanding phase uses priori information about the sketch to support identification of symbols by excluding any possible invalid symbols. 
%the basic symbols are identified in consistent with the whole sketch. \\
%from the user information from the users converts the strokes the user draw into segments which is recognized into basic symbols that can be easily manipulated. Finally, the basic symbols are interpreted in consistent with the whole sketch. \\
This paper introduces a sketch recognition system. The presented system uses particle swarm algorithm(PSO) to optimially segment users strokes. Users can draw symbols using any number of strokes in any order they like. The system segments stroke user draws then cluster symbols together then passes them to the SVM symbol recognizier. The experiments in section \ref{sec:Experiments} shows that the proposed system improves the final symbol recognition accurracy.

The remaining of this papers is organized as following; section \ref{Sec:RelatedWork} provides a brief literature review.  Section \ref{PSO} explains the general particle swarm algorithm which is used in the presented system. The proposed system block diagram along with a detailed system discribtion is presented in Section \ref{Sysdisc}.  The experiments and results achieved are explained in Section \ref{sec:Experiments}. Finally, section \ref{ConclusionandFutureWork} states the conclusion and future work. 
%The next section will relate previous work in the field. The remaining of the paper will be as following; Section \ref{PSO} will explain the general particle swarm algorithm. Section \ref{Sysdisc} will describe the system block diagram and then section \ref{Prepross} will explain the preprocessing step, section \ref{seg} will describe the segmentation algorithm in details. Finally, section \ref{sec:Experiments} will describe the experiments that was preformed.\\
\section{Literature Review}
\label{Sec:RelatedWork}
%%%What is sketch recognitions. Why we need sketch recognition systems. 
The field of symbol recognition has gained interest in the last few years. A wide variety of techniques were used either on segmentation process or symbol recognition.
%To segment strokes
%In order to segment strokes, \cite{meanshift10} introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on this area.  %\citeauthor{HeloiseBeautification} used a dynamic programming algorithm to solve the segmentation problem \cite{HeloiseBeautification}. 
A hybrid algorithm was introduced in \cite{earlyprocess} where they generated different sets of segmentation based on both curvature and speed dominant points, this is followed by choosing a segmentation with the least error from generated hybrid set. Their system has a draw back as it only recognizes simple specific shapes with a set of low level recognizers. \cite{meanshift10} introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on the computed feature area. Their system achieved good accuracy in simple shapes (square, ellipse,...) but did not perform well in more complex shapes\cite{meanshift10,Paleosketch08}. \citeauthor{Paleosketch08} \cite{Paleosketch08} introduced a set of low level recognizers that were reported to achieve 98\%  accuracy but their system simillar to all low level recognizers only identify a small set of simple shapes.

Many different approaches have been investigated to achieve final recognition of symbols. Graph searching and template matching where used in \cite{GraphBased07} where a graph of the drawn symbol is generated and matched with a stored set of graphs templates to reach a classification decision. These systems may generate good performance but they are very computationally expensive. Geometrical and spatial descriptive language was used in \cite{SketchRead2007,Ladder30} where an architecture similar to compiler was used to recognize symbols. %%SVM and HMM classification was implemented in\cite{SVM300,HMM53} to correctly classify symbols.

Genetic algorithm was used by \cite{CruveDivisionSwarm} to optimally divide digital curves into lines and curves. \citeauthor{CruveDivisionSwarm} uses digital curves scanned from paper as input to the system and did not take advantage of the curvature or local geometric properties of the digital curve. \citeauthor{PolygonApproximationPSO} \cite{PolygonApproximationPSO} used PSO to convert digital curves into polygons, our system adoptes \citeauthor{PolygonApproximationPSO} system but improves it by adding curvature and other local information while segmenting strokes to achieve better segmentations. The next section presents the general particle swarm algorithm. The proposed swarm segmentation algorithms are peresented in section \ref{subsubsec:Discreteparticleswarmalgorithm}
% Other techniques as  [ ] was used  in []. \\  
%In the next section the particle swarm algorithm will be explained. After that the system block description and detailed algorithms will be described. We will then proceed to describe the experiments preformed and the achieved results.  
%\subsection{Sketch Recognitions systems}
%Details of Sketch systems. 
%Related work and research in this area. 
%Writing on the pros and cons of current systems.  
%Write on the difficulties of the sketch recognition systems. 
  %Some tried to detect dominant or corner points based on data generated from speed   \cite{earlyprocess} and mostly used curvature of the stroke drawn \cite{earlyprocess,meanshift10}. 
%Particle swarm algorithms was used in \cite{} to fit a set of points into ellipse or higher order curves.
%The first set of features introduced to be used on gesture recognition was introduced by Rubine in \cite{gestureexample12}. later,  Spatial relations and features where used in  \cite{cognitivesketch18,SRGraph57}  . \citeauthor{GeometryAndDomain102} introduced ink density feature which they used.  The use of Zernike moments  was used  as the only features to the SVM classifiers in \cite{HeloiseBeautification}.  \cite{SketchRead2007} used a set of features varied from relative spatial features and some compositional features.  \\
%After segmentation researchers varies on the methods in which to continue.
\section{Particle Swarm Algorithm}
\label{PSO}
%What is particle swarm algorithm and how it was used in related researches. 
The main idea of  Particle Swarm Algorithm (PSO) is to represent each agent with a particle from the solution space. Each agent moves the particle with a direction and velocity based on equation \ref{eq:Swarm}.
 %Equation [\ref{eq:Swarm}] shows how velocity and direction of each particle are computed
 \begin{equation}
v_{ij}  = v_{ij}  + c_1 r_1 (lbest_{ij}  - p_{ij} ) + c_2 r_2 (gbest_{ij}  - p_{ij} )
\label{eq:Swarm}
\end{equation}
 where $r_1$ \& $r_2$ are random variables and $c_1$ \& $c_2$ are the swarm system variables.
\begin{equation}
%\[
p_{ij}=p_{ij}+v_{ij},
%\
\end{equation}
where $p_{ij}$ represent the $jth$ particle in the $ith$ agent and $v_{ij}$ is the velocity of the $jth$ particle in the $ith$ agent. Each iteration the global best $g_{best}$ particle and the agent local best $l_{best}$ particle are evaluated based on the maximum fitness functions of all particles. The solution is found after acheiving a specific number of iteration or after an error threshold is achieved.
The following equation \ref{eq:descrite}  
\begin{equation}
   P(i)\Leftarrow 
\{
\begin{array}{c} 
1 \quad \quad if\quad r_{3}>t  \\

0 \quad \quad if\quad r_{3}<t 
\label{eq:descrite}
\end{array}\}
\end{equation}
 (where $t$ is a threshold and $r_{3}$ is a random variable), is used in order to change the general swarm algorithm into binary particle which handles particle values of either $0$ or $1$.

\section{System Description}
\label{Sysdisc}
%The stroke is the path of points from the instant the pen is down till it is up.
Figure. \ref{fig:Blockdiagram} shows that the main system blocks are 1) preprocessing, 2) segmentation, 3) clustering, 4) feature extraction, and finally 5) classification.

 %Due to sloppiness of users and hardware glitches the system preprocess the data before further processing. 
The preprocessing is responsible of capturing the input data and removing the noise from it. The system then proceeds to estimate a set of possible dominant points to help in the segmentation process. % The time differences, direction, speed and curvature of all points of the stroke is computed along with some other geometrical data. The system then proceed to estimate a set of possible dominant points to help in segmentation process.  \\
  %then passes through the segmentation process which will try to

@@

 %the stroke first If the ellipse detection fails the stroke is passed to the segmentation algorithms which will pass it to the two PSO algorithms described below the segmentation with the minimum error will be the chosen segmentation.  The segmentation is then added to the set of un-recognized segments in the system. \\% This part is repeated for each stroke. 
 The clustering algorithms starts to group segments together after the segmentation step. The system let the user draws the symbol by using any number of strokes, a set of unrecognized segments is passed to the clustering algorithm to generate a symbol and compute a feature vector for it. The system uses a hybrid set of statistical, geometric and spatial features. The final step is a SVM classifier that will use the features computed to classify the segments into one of the previously trained classes. %  Or to determine the symbol of the given segments from the set of preciously trained symbols. 
 %with a symbol from the training set.% The system compute composite set of features some are statistical other are spatial features based on the type primitives. Finally, the strokes are classified into the corresponding classes.% the classifier identifies the strokes into a symbol from the set of known symbols.\\

\begin{figure}[]
%\begin{minipage}[b]{0.5\linewidth}
	\centering
		\includegraphics[scale=0.85]{images/block2.pdf}%Blockdiagram.pdf}
	\caption{The system Block Diagram.} The figure shows the main system blocks. Preprocessing, segmentation, clustering,  features extraction and classification.

	\label{fig:Blockdiagram}
%	\end{minipage}
\end{figure}

\subsection{Preprocessing}
\label{Prepross}
%Some data is extracted from the points in the stroke.
 It is noted that as the user draws a shape the pen will slow down near corners and picks up speed when drawing a straight lines. Therefore, the speed and time difference data were widely used in sketch understanding systems \cite{earlyprocess}.  The direction and curvature data are used to determine the points with high angular changes along the path of points \cite{meanshift10}.

@@

 %system compute speed, distance data for all points in the stroke. Curvature is computed using estimation used in [] where direction computed is angle between two lines. Curvature is 
 %$\Delta d/  \Delta S $ where d is the difference direction of point and s is difference in distance between points. 
 All these calculations are performed in real time while the user draws the strokes. The complexity of computation is $O(n)$ where $n$ is number of points. \\
 Figure \ref{fig:speed2} shows the computed data for the stroke drawn in fig. \ref{fig:orignalStroke}. If you look at the data curves you will see that dominant points are characterized by lower speed values and higher curvature and direction values. \\% the lower speed points correspond to vertex and dominant point's location. The higher the direction and curvature data correspond to location with higher change in curvature witch promote the location for vertices. \\
After the system computes all the speed, time difference and curvature data it proceeds to detect the points with low velocity and high curvatures. Using simple differentiations to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted a process presented by \cite{earlyprocess}, where the mean
%So instead the following procedure was used, firstly the mean of the curve is used as threshold for lower values then the high values are further investigated to find the extreme points. Hence, the mean 
of the curve is calculated and used as threshold. This threshold is used to separate the curve into regions; each region is defined as the part between two intersection points of the threshold line and the curve. Regions higher than the threshold value are processed to find the maximum of each region. The points that correspond to those maximum values are labeled as \textit{possible dominant points}.   \\
   %If we tried to differentiate the curve the result will be false threshold values we divide the data into regions of data higher or lower that the threshold. This will let us only look for high data only. The maximum of each region is then computed and reported as a possible vertex point. 
\begin{figure}[]
	\centering
		\includegraphics[scale=0.3]{images/orignalStroke.pdf}
	\caption{Stroke Sample} An example of the stroke drawn in the system.
	\label{fig:orignalStroke}
\end{figure}
The system repeat this process to curvature, time difference and speed curves. All the points labeled as possible dominant points are saved into a single array. They are used as input to guide the next stage of segmentation. 
\begin{figure*}[]
%\begin{minipage}[b]{0.9\linewidth}
	\centering

		\subfigure[ The Speed of data] {\includegraphics[scale=0.25]{images/speed2.pdf}}	
	%	\subfigure[ Direction ] {\includegraphics[scale=0.15]{images/direction2.pdf}}
					%\end{minipage}
					\hspace{20 mm}
					%\begin{minipage}[b]{0.5\linewidth}
			\subfigure[The Curvature] {\includegraphics[scale=0.29]{images/curvature2.pdf}}
	%		\subfigure[ Time Difference] {\includegraphics[scale=0.15]{images/time2.pdf}}
%			\end{minipage}
	\caption{Data Curves} The curve in a) shows the speed of the points in stroke in fig. \ref{fig:orignalStroke}. b) Shows the curvature curve for the same stroke.  %{The data of the stroke}

@@

After computing the primary data the system tries to segment the stroke into a set of primitives. The segmentation algorithm first tries to detect if the stroke is an ellipse. If the stroke proved to be an ellipse then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed into two particle swarm algorithms that will divide the stroke to either lines or lines and curves. The algorithms takes the stroke points along with the possible dominant points computed then produce a set of dominant points which are connected with either lines or curves (see fig. \ref{fig:Blockdiagram}). The next section will describe the ellipse detection algorithm and both the two particle swarm algorithms used to divide the stroke.\\

@@

\subsubsection{Ellipse Detection} 

@@

The process starts with computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation see equation \ref{eq:circleFit}.

@@

\end{equation} where N is number of points in the stroke, $a,b$ are the length of ellipse axes, $x_0$ \& $y_0$ are the coordinates of the center point, $x_i$ \& $y_i$ are the coordinates of point $i$ in the stroke. After only few loops the error and a confidence values  are computed to check if the stroke can be labeled as ellipse or not.\\ %is computed to check if stroke is an ellipse. \\
\subsubsection{Discrete particle swarm algorithm}
\label{subsubsec:Discreteparticleswarmalgorithm}

@@

\subsubsection{Problem definition}

@@

let's define the arc $\widehat{x_ix_j}$ as the consecutive set of points from point$x_i$ to point $x_{j}$ as in $x_i,x_{i+1} \cdots,x_j$. The line $\overline{x_i x_j} $ as the straight line connecting point $x_i$ to point $x_j$. The approximation error is computed by the equation \ref{eq:ErrorSwarm1}

@@

\end{equation} where M is the number of dominant points in this solution.  The error $e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$.  \\
The fitness is computed using equation \ref{eq:fitnessSwarm1}\begin{equation}

@@

\end{equation} where $N$ is the number of points in the stroke, $D$ is the number of point in the solution that was previously labeled as a possible dominant point and $E$ is the computed error and $\varepsilon$ is the error threshold.  Notice that when the error is larger than the threshold the fitness is given a -ve value to lower the value of solution otherwise the system will favor the lower number of vertices.\\ %we will want to lower the number of vertices. \\

@@

%Alg2: o
The second algorithm has the same problem formulation but different fitness and error functions. It was previously introduced in  \cite{CruveDivisionSwarm} but a genetic programming was used as the optimizing algorithm.  The error of both circle and line estimation are computed for each segment, the approximation with the lower error value will be the chosen approximation of this segment. The sum of the approximation error of all segments is the total error of the particle.  The error is computed by equation \ref{eq:errorSwarm2} \begin{equation}

@@

\end{equation}where $M$ is the number of segments in the solution, $D_i$ is the minimum approximation error of curve and line approximations $min(d_c,d_l)$ as computed by \cite{CruveDivisionSwarm}.  The fitness is computed by the equation \ref{eq:fitnessSwarm2} \begin{equation}

@@

\end{equation} where $E$ is the error and $M$ is number of segments and $k$ is a parameter tweaked to get minimum number of segments. \\

@@

\citeauthor{PolygonApproximationPSO} \cite{PolygonApproximationPSO} used a merge and divide algorithm after each loop of the swarm system to refine the solution but this system is introducing another enhancement method. After each loop of the swarm algorithm, each particle is refined using the following procedure. For each particle %loops on the set of selected dominant points to enhance the solution.
each dominant point is checked to find if it was labeled before as a \textit{possible dominant point} (computed as in section \ref{Prepross}), if it is not labeled the point is moved to the nearest labeled point. After that the particles are tested to make sure that the distance between every two successive dominate point is larger than the constant $min_D$. If two points are found that they are nearer than $min_D$ then based on the decrease in the error of the solution one of the points will either be removed or moved to the nearest possible dominate point. 

@@

After the user draws all strokes of the symbol he has to wait 10 seconds or press finish button beside the drawing area. The set of unrecognized strokes is grouped together along with their segmentation as input to the feature extraction process. The system uses a composite set of features include Rubine feature set \cite{gestureexample12}, Zernike moments \cite{HeloiseBeautification}, ink density \cite{GeometryAndDomain102} and some structural and spatial information like number of perpendiculars lines , number of parallel lines and types of primitives in each symbol. After computing the features the symbol is introduced to the classifier. The system used a support vector machine (SVM) classifier with Gaussian kernel (RBF kernel) \cite{libsvm}.\\%Some features like number of parallel, perpendicular and intersecting lines are computed from the segmentation other like density, centroid and Zernike moments are computed from the raw stroke points. 
